---
title: "Introduction to circaTime"
author: "Alperen Uysal"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to circaTime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libs, message = FALSE}
library(circaTime)
library(SummarizedExperiment)
```

## Why circaTime

Tens of thousands of public RNA-seq samples have no recorded collection time, yet
circadian phase — the time of sample collection relative to the internal
clock — can often be recovered computationally from expression data alone.
Several published methods do this, but each lives in its own repository, takes its
own input format, and reports its own notion of confidence (usually none at all).
`circaTime` wraps published phase-inference methods behind a single
`SummarizedExperiment`-based interface, adds a benchmarking harness so methods can
be compared on the same labelled ground truth, and adds a calibrated uncertainty
layer — a confidence interval and a quality flag — rather than a bare point
estimate.

This vignette walks through the whole pipeline: a raw `SummarizedExperiment`, the
call that adds phase estimates to it, how to check those estimates against known
ground truth, what the added confidence columns mean, and how a phased
`SummarizedExperiment` hands off to downstream rhythmicity analysis.

## The data

The example data bundled with the package is a real subset of GSE54650 (Zhang et
al. 2014, *PNAS*): mouse liver, 24 samples spanning one full circadian cycle at
2-hour resolution, full gene set. Real published data was used here specifically so
every number in this vignette is a real result, not a simulation.

```{r load-data}
se <- readRDS(system.file("extdata", "example_se.rds", package = "circaTime"))
se
```

`colData(se)` happens to carry a `phase` column here because GSE54650's collection
time is known — that is what lets this vignette also demonstrate benchmarking below.
Most real-world input to `estimatePhase()` will *not* have this column; that is
precisely the value the package provides.

```{r show-truth}
head(colData(se))
```

## From raw to phased: `estimatePhase()`

`estimatePhase()` is the package's main entry point. It requires a `tissue` (there
is no default — applying a model trained on one tissue to another silently is a
bug, not a convenience) and looks up a matching pre-trained reference model.
`availableReferenceModels()` lists what is currently bundled:

```{r avail}
availableReferenceModels()
```

```{r estimate}
phased <- estimatePhase(se, method = "timetable", tissue = "Liv")
colData(phased)[, c("phase", "phase_lo", "phase_hi", "clock_amp", "phase_flag")]
```

Five columns are added: `phase` (the point estimate, hours in `[0, 24)`),
`phase_lo`/`phase_hi` (a confidence interval — see below), `clock_amp` (core-clock
amplitude — see below), and `phase_flag` (a quality flag: `"ok"`,
`"low_amplitude"`, `"gene_coverage"`, or `"failed"`). `estimatePhase()` never
hard-errors on a difficult individual sample; a sample that cannot be scored gets
`phase = NA` and an informative flag instead. Only malformed input (an unrecognised
tissue/species/method combination, or an assay with no gene identifiers) is a hard
error.

## Not sure which engine to pick? `method = "consensus"`

No engine wins everywhere — across the package's full 12-tissue GSE54650 benchmark
(`benchmarks/run_consensus_benchmark.R`), three different engines each win in at
least three tissues. Rather than forcing that choice on you,
`method = "consensus"` runs every engine bundled for that tissue (for mouse
microarray: Molecular Timetable and tauFisher) and combines their estimates on the
circle:

The two chunks below are shown but not executed here: `method = "consensus"`
combines `timetable` with `tauFisher`, and `tauFisher` is a GitHub-only package
(`Suggests`, not installed by Bioconductor's or r-universe's build system). Install
it locally (`remotes::install_github("micnngo/tauFisher")`) and re-run these chunks
to see the real numbers.

```{r consensus, eval = FALSE}
consensus <- estimatePhase(se, method = "consensus", tissue = "Liv")
colData(consensus)[, c("phase", "phase_lo", "phase_hi", "phase_flag")]
```

The combination is not a black box. Everything that went into it is returned:

```{r consensus-report, eval = FALSE}
report <- S4Vectors::metadata(consensus)$circaTime_consensus
report$stored_weights      # the trained prior, per engine
report$per_engine_phase    # what each engine said, per sample
round(report$concentration, 3)  # how strongly the engines agreed, in [0, 1]
```

Two weighting schemes are available. The default, `consensus_weights = "accuracy"`,
weights each engine by `1/cvMAE` using that engine's own held-out error **on this
specific tissue** (visible as `cv_mae` in `availableReferenceModels()`), so an
engine that does well on liver but poorly on brainstem is weighted differently in
each. `consensus_weights = "equal"` is a plain circular mean, useful as the honest
baseline for whether the accuracy weighting is buying anything. These are two
separately trained models, not one model with a switch — each carries its own
confidence interval, measured under its own weighting.

If one engine's genes are missing from your data, its weight drops to zero and the
rest are renormalised, rather than it contributing an unusable estimate; the result
is only flagged `"gene_coverage"` if no engine had coverage.

Benchmarked with real donor-split cross-validation across all 12 GSE54650 tissues
(`benchmarks/run_consensus_benchmark.R`): pooled mean circular MAE was 0.723h
(accuracy-weighted consensus), 0.726h (equal-weighted), 0.738h (Molecular
Timetable alone), 0.804h (tauFisher alone) — both consensus schemes beat
committing to either single engine everywhere, though the two weighting schemes
are close enough to each other (0.723h vs. 0.726h) that this is not a strong
argument for tissue-aware weighting specifically over the simpler equal-weighted
baseline. See `notes/decisions.md` Section 14 for the full accounting, including
the oracle (best-single-engine-per-tissue, not achievable in practice) comparison.

## Gene identifiers across ID types and species

Real input data does not always arrive as mouse gene symbols. `standardizeGeneIds()`
auto-detects whether `rownames(se)` are gene symbols, Ensembl IDs, Entrez IDs, or
RefSeq IDs (per gene, so mixed input is fine), and maps everything to gene symbol —
the identifier convention every other function in this package uses:

```{r standardize, eval=requireNamespace("org.Mm.eg.db", quietly = TRUE)}
mixedIds <- SummarizedExperiment(
  assays = list(logExpr = matrix(rnorm(9), nrow = 3,
    dimnames = list(c("Arntl", "ENSMUSG00000029238", "11461"), c("s1", "s2", "s3"))))
)
mixedIds <- standardizeGeneIds(mixedIds, species = "Mmusculus")
rownames(mixedIds) # Arntl (already a symbol), Clock (Ensembl), Actb (Entrez)
```

Two layers do the work. A dependency-free layer runs first: it strips Ensembl
version suffixes and resolves a small, hand-curated table of common circadian
clock-gene synonyms (e.g. `BMAL1`/`MOP3` → `ARNTL`, `DEC1` → `BHLHE40`, `CHRONO`/
`C1orf51` → `CIART`) — curated by hand from HGNC/MGI "previous symbol" and "alias"
fields for the core clock genes this package's bundled models and QC panel
actually use, not pulled programmatically from a live database (so it is small,
auditable in `R/gene-mapping.R`, and does not need a network call). Anything the
free layer cannot resolve — a real Ensembl/Entrez/RefSeq ID, or a symbol outside
that curated set — falls through to `AnnotationDbi::mapIds()` against
`org.Mm.eg.db`/`org.Hs.eg.db` if installed (`Suggests`, not required for the common
already-symbol case).

Cross-*species* mapping is a separate, harder problem from cross-*ID-type* mapping
within one species, and `mapOrthologs()` handles it via the `babelgene` package's
curated one-to-one mouse↔human ortholog tables — a gene with no unambiguous
ortholog is dropped (`NA`), not guessed at:

```{r orthologs, eval=requireNamespace("babelgene", quietly = TRUE)}
mapOrthologs(c("Arntl", "Clock", "Per1"), fromSpecies = "Mmusculus", toSpecies = "Hsapiens")
```

`estimatePhase()` wires this in directly via `data_species`: set it when `se`'s
genes are in a different species' convention than the bundled model (currently
mouse-only) you are calling against, e.g. human input data against the bundled
mouse model. `map_orthologs = TRUE` is the default whenever the two differ, and a
mapping summary is recorded at `metadata(se)$circaTime_ortholog_mapping`, the same
way `circaTime_gene_mapping` already records the within-species alias/ID-type
report.

## How good is this? Benchmarking against known ground truth

Because this example happens to carry real ground truth, we can check point-estimate
accuracy directly with `benchmarkPhase()`, which cross-validates a method with
held-out folds split by donor (never by sample, so no individual's data leaks across
train/test):

```{r benchmark}
bm <- benchmarkPhase(se, truth_col = "phase", methods = "timetable", nFolds = 5, seed = 1)
bm$summary
```

`circular_mae` is mean absolute error on a 24-hour circle — computed as
`pmin(abs(d), 24 - abs(d))`, since 23.5h and 0.5h are one hour apart, not 23.
Ordinary (non-circular) statistics are not meaningful on phase and are never used
anywhere in this package. This result matches what was reported for liver in the
package's own benchmark suite (`circular_mae` $\approx$ 0.98h across the full
12-tissue GSE54650 atlas; see the package's `benchmarks/results/` for the complete,
reproducible table).

`benchmarkPhase()` answers "how accurate is this engine on held-out samples from
*this* dataset?" A different, complementary question is transfer: "if I train on
one dataset/platform, how accurate is it on a *different* one?" — e.g. training on
microarray data and scoring RNA-seq samples of the same tissue. `transferPhase()`
answers that, reusing the same engine registry and circular-error metrics as
`benchmarkPhase()`, just with the train/test split fixed by which dataset a sample
came from instead of drawn as cross-validation folds:

```{r transfer, eval = FALSE}
seMicroarray <- circaTimeData::circaTimeData("GSE54650") # same study, microarray
seRnaseq <- circaTimeData::circaTimeData("GSE54651")     # same study, RNA-seq
liverMicro <- seMicroarray[, SummarizedExperiment::colData(seMicroarray)$tissue == "Liv"]
liverRnaseq <- seRnaseq[, SummarizedExperiment::colData(seRnaseq)$tissue == "Liv"]

tb <- transferPhase(liverMicro, liverRnaseq, truth_col = "phase", methods = "timetable")
tb$summary
```

## Confidence intervals and quality flags

`phase_lo`/`phase_hi` are not a bootstrap refit on every call (that would be far too
slow for some engines) — they are built from a pool of real, held-out prediction
residuals collected once when each reference model was trained, added directly to
the point estimate. This was checked directly against real held-out data: a stated
80% interval that used only training-resampling variance (no residual pool) covered
the true phase only about 48% of the time — badly overconfident. Adding the
residual-pool mechanism brought empirical coverage in line with (in fact slightly
above) the nominal level at every level tested. An uncalibrated interval is worse
than none, so this check was treated as a hard requirement, not an afterthought.

`phase_credible_lo`/`phase_credible_hi` is a second, complementary interval: a
Bayesian (von Mises) credible interval built from the *same* residual pool as
`phase_lo`/`phase_hi`, via `bayesianPhasePosterior()`. Where the empirical
interval is the raw quantile spread of the observed residuals, the Bayesian one
fits a smooth parametric distribution to the pool's concentration and returns
its analytic interval - useful when a caller needs more than interval endpoints
(e.g. `bayesianPhasePosterior()` also returns a full posterior density,
enabling statements like "what is the probability the true phase is within 2h
of the point estimate"). The two intervals are expected to be close, not
identical: on real held-out GSE54650 predictions, pooled empirical coverage at
nominal 80% was 88.9% (Bayesian) vs. 87.2% (bootstrap), with mean widths 2.72h
vs. 2.78h (`benchmarks/run_bayesian_ci_benchmark.R`). `kappa`/`concentration`
are recorded in `metadata(se)$circaTime_bayesian`.

```{r bayesian}
metadata(phased)$circaTime_bayesian
colData(phased)[, c("phase", "phase_lo", "phase_hi",
                    "phase_credible_lo", "phase_credible_hi")]
```

`clock_amp` (`phase_flag == "low_amplitude"` when it falls below threshold) has two
implementations, selected with `qc_method`:

- `qc_method = "legacy"` (default): the Euclidean norm of a sample's first two
  principal components over a **fixed** 16-gene core clock panel, projected onto
  the reference cohort's own PCA basis (works even for a single-sample call).
  **Honest caveat**: on the two real labelled datasets it was tested against (the
  healthy-tissue GSE54650 atlas used above, and a young-vs-aged mouse stem cell
  dataset with real circadian-time ground truth), low-amplitude samples did not
  show a statistically significant error difference from high-amplitude ones.
  Core clock *gene* expression is reported in the aging literature to stay close
  to unaltered while *downstream* oscillators desynchronise — so a fixed
  core-clock panel may simply be looking at the wrong genes.
- `qc_method = "dynamic"` (`clockCoherence()`): selects the informative gene
  panel **from the input dataset itself** — each gene's cosinor fit is tested
  against a phase-permutation null, panel = genes clearing a BH-FDR cutoff —
  rather than trusting one fixed panel across every dataset. This needs a real
  cohort in the call (`qc_min_samples`, default 10) since the panel is fit from
  it; below that it silently falls back to `"legacy"`
  (`metadata(se)$circaTime_qc$method_used` records which ran). On the same
  young-vs-aged dataset where the legacy flag scored p=0.20, the dynamic score
  reaches a selection-corrected p=0.0185 — see
  `benchmarks/run_dynamic_qc_benchmark.R` and `notes/decisions.md` Section 15
  for the full validation and why coherence is a *self-consistency* score (it is
  scored against `estimatePhase()`'s own phase estimate, not independent ground
  truth) rather than a general-purpose sample-quality metric.

```{r dynamic-qc}
se_dyn <- estimatePhase(se, method = "timetable", tissue = "Liv",
                        qc_method = "dynamic")
table(colData(se_dyn)$phase_flag)
metadata(se_dyn)$circaTime_qc[c("method_used", "threshold", "panelSize")]
```

## Handing off to downstream rhythmicity analysis

A phased `SummarizedExperiment` is meant to be handed straight to existing
rhythmicity-detection tools rather than re-implementing what they already do. The
pattern below (illustrative — `limorhyde` and `dryR` are not dependencies of
`circaTime` and are not installed or executed here) shows the shape of that
handoff: `circaTime` supplies `colData(se)$phase` as a covariate, the downstream
tool takes it from there.

```{r downstream, eval = FALSE}
# Illustrative only - not run as part of this vignette.
library(limorhyde)
colData(phased)$time_var <- limorhyde(colData(phased)$phase, colnames(colData(phased)))
# ... proceed with limorhyde's own differential-rhythmicity workflow, e.g. via limma
```

## Is circadian time confounding my design?

Most users of this package are not circadian researchers. They have a differential
expression design — cases versus controls, treated versus untreated — and a
reasonable worry: the samples' collection times were never recorded, so could an
unmodelled daily rhythm be driving what looks like a condition effect?

That worry is only justified under a specific condition. Circadian phase biases a
contrast when it is **unbalanced across the groups being compared**. If collection
time varies randomly *within* every group, rhythmic genes inflate residual variance
but leave the condition coefficient unbiased. If the cases happened to be collected
in the morning and the controls in the afternoon, then every rhythmic gene carries an
apparent condition effect that is really a clock effect.

`phaseConfounding()` tests exactly that: it asks whether phase is distributed
unevenly across each term of a design formula.

```{r confounding}
# A deliberately confounded design: cases collected around ZT6, controls around ZT14.
set.seed(1)
n <- 20
confPhase <- c(rnorm(10, mean = 6, sd = 1), rnorm(10, mean = 14, sd = 1))
confMat <- matrix(rnorm(100 * n), nrow = 100,
                  dimnames = list(paste0("g", seq_len(100)), paste0("s", seq_len(n))))
conf <- SummarizedExperiment(assays = list(logExpr = confMat))
colData(conf)$phase <- wrapPhase(confPhase)
colData(conf)$condition <- rep(c("case", "control"), each = 10)

phaseConfounding(conf, ~ condition, nPerm = 999, seed = 1)
```

The small p-value says the imbalance is real, and `effect_size` puts it in
interpretable units: the group mean phases sit roughly eight hours apart. Compare
that with a design where the same samples were collected across the whole day in both
arms:

```{r confounding-balanced}
balanced <- conf
colData(balanced)$phase <- rep(seq(0, 22, by = 2), length.out = n)

phaseConfounding(balanced, ~ condition, nPerm = 999, seed = 1)
```

Here phase is spread evenly through both groups, and the test says so.

Three things are worth being precise about:

- **Every comparison is circular.** Groups centred on ZT23 and ZT1 are two hours
  apart, not twenty-two, and both the test statistic and `effect_size` treat them
  that way. Applying an ordinary ANOVA or t-test to phase values gets this wrong.
- **The p-value comes from permutation**, not from the usual parametric circular
  approximations, which assume von Mises phases and equal concentration across
  groups — assumptions inferred phases routinely violate.
- **Nothing is corrected for you.** A flagged design is information, not an
  instruction: whether to add phase as a covariate, drop the affected samples, or
  simply report the caveat is a modelling decision that belongs to the analyst.

If `se` does not already carry a `phase` column, pass `method` and `tissue` and
`phaseConfounding()` will call `estimatePhase()` first. Samples whose phase estimate
was flagged unreliable (`low_amplitude`, `gene_coverage`, `failed`) are excluded from
the test rather than allowed to add noise to it; the number dropped is reported in
`attr(result, "nDropped")`.

## What isn't covered yet

This is an early snapshot, and it is worth being explicit about the edges:

- Only mouse (`species = "Mmusculus"`) tissues from GSE54650 are currently bundled;
  see `availableReferenceModels()` for the exact list. A larger, more
  tissue/species-diverse reference cohort lives in the separate companion
  `circaTimeData` ExperimentHub package rather than being bundled directly in
  `circaTime` itself.
- `standardizeGeneIds()`/`mapOrthologs()`/`estimatePhase(data_species = )` (above)
  cover ID-type translation and one-to-one mouse↔human ortholog mapping, but there
  is still no *bundled reference model* trained on human data to map onto — an
  ortholog-mapped human sample is currently scored against the mouse model, which
  is unvalidated cross-species transfer, not a substitute for a real human cohort.
- `phaseConfounding()` tests each design term separately. Interaction terms are
  skipped, and a full multivariable circular regression — which would let you ask
  whether phase is imbalanced across one term *after adjusting for* another — is not
  implemented.
- Cross-tissue and cross-platform (microarray vs. RNA-seq vs. pseudobulk) transfer
  has not yet been measured systematically; applying a model trained on one tissue's
  platform to a very different one should be treated as unvalidated until it has
  been.

## Session info

```{r session-info}
sessionInfo()
```
